\documentclass{siamart}

\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{graphicx}

\newtheorem{remark}{Remark}

\lstdefinestyle{python}{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!70!black}\bfseries,
  commentstyle=\color{gray},
  stringstyle=\color{orange!80!black},
  showstringspaces=false,
  frame=single,
  framesep=4pt,
  xleftmargin=8pt,
  xrightmargin=8pt,
  belowskip=6pt,
  aboveskip=6pt,
}

\title{Krylov Methods for the Markowitz Portfolio}

\author{Thomas Schmelzer%
  \thanks{Jebel Quant (\email{thomas@jqr.ae}).}
  \and
  Raphael Hauser%
  \thanks{Mathematical Institute, University of Oxford.}
  \and
  Stephen Boyd%
  \thanks{Department of Electrical Engineering, Stanford University.}
}

\headers{Krylov Methods for the Markowitz Portfolio}{T.~Schmelzer, R.~Hauser, and S.~Boyd}

\begin{document}

\maketitle

\begin{abstract}
TODO
\end{abstract}

\section{Introduction}

The companion paper \cite{minvar_paper} showed that the long-only minimum variance portfolio
reduces to a symmetric indefinite KKT system solvable by MINRES without ever forming the
sample covariance matrix, yielding speedups of 354--644$\times$ over CVXPY on standard benchmarks.
The present paper extends this framework to the full Markowitz mean-variance problem.

Three generalisations are required: (i) the single budget constraint $\mathbf{1}^\top w = 1$ is
replaced by a general equality constraint matrix $A$; (ii) general linear inequalities
$C^\top w \leq d$ are handled by extending the active set loop; and (iii) a linear return term
$\rho\,\mu^\top w$ is added to the objective. In each case the KKT saddle-point structure is
preserved and MINRES applies without modification.

\section{Problem Formulation}\label{sec:problem}

Let $X \in \mathbb{R}^{T \times n}$ be the matrix of daily returns for $n$ assets over $T$ trading
days, $\mu \in \mathbb{R}^n$ a vector of expected returns, and $\rho \geq 0$ a risk-aversion
parameter. The general mean-variance problem is:
\[
\min_{w} \; \|Xw\|^2 - \rho\,\mu^\top w
\quad \text{subject to} \quad
A^\top w = b, \quad C^\top w \leq d
\]
where $A \in \mathbb{R}^{n \times m}$, $b \in \mathbb{R}^m$ encode $m$ linear equality constraints
(at minimum the budget constraint, and optionally sector limits, factor-exposure constraints, etc.),
and $C \in \mathbb{R}^{n \times p}$, $d \in \mathbb{R}^p$ encode $p$ linear inequality constraints
(the long-only requirement $w \geq 0$ being the special case $C = -I$, $d = \mathbf{0}$).
Setting $\rho = 0$ and $A = \mathbf{1}$, $b = 1$, $C = -I$, $d = \mathbf{0}$ recovers the
minimum variance problem of \cite{minvar_paper}.

\section{The KKT System}\label{sec:kkt}

\subsection{Equality-constrained subproblem}

Ignoring the inequality constraints, the Lagrangian is:
\[
\mathcal{L}(w, \lambda) = \|Xw\|^2 - \rho\,\mu^\top w - \lambda^\top(A^\top w - b)
\]
The stationarity condition $\nabla_w \mathcal{L} = 0$ gives $2X^\top X w - A\lambda = \rho\mu$.
Combined with $A^\top w = b$, this yields the $(n+m)\times(n+m)$ saddle-point system:
\[
\begin{pmatrix} 2X^\top X & A \\ A^\top & 0 \end{pmatrix}
\begin{pmatrix} w \\ \lambda \end{pmatrix}
=
\begin{pmatrix} \rho\mu \\ b \end{pmatrix}.
\]
The matrix is \textbf{symmetric but indefinite}; MINRES \cite{paige1975} handles this directly.
Setting $\rho = 0$ recovers the minimum variance KKT system of \cite{minvar_paper} with the
right-hand side $(\mathbf{0},\, b)$. The constraint-eliminated CG formulation generalises
analogously: the null-space basis $P$ becomes an orthonormal basis for $\ker(A^\top)$, of
dimension $n - m$ when $A$ has full column rank.

\subsection{Inequality constraints and the active set method}

General linear inequalities $C^\top w \leq d$ are handled by extending the active set loop.
At each outer iteration, the currently active inequalities (those binding at the current
solution) are appended as additional columns of $A$; the KKT saddle-point system is solved on
the resulting enlarged equality-constrained subproblem. Constraints that become inactive are
dropped and the solve is repeated. The process terminates when no active-set change occurs.
The long-only constraint $w \geq 0$ is the special case $C = -I$, $d = \mathbf{0}$, already
present in \cite{minvar_paper}.

\subsection{Efficient frontier}

Varying $\rho$ from $0$ to $\infty$ traces the efficient frontier: $\rho = 0$ gives the
minimum variance portfolio; large $\rho$ gives the maximum return portfolio. Each point on the
frontier requires one active set solve of the KKT system above, with a different right-hand
side $(\rho\mu,\, b)$ but the same saddle-point matrix. Warm-starting from the previous
$\rho$ reduces the active set iterations.

\section{Computational Methods}\label{sec:solvers}

TODO

\section{Numerical Results}\label{sec:results}

TODO

\section{Conclusion}

TODO

\section*{Acknowledgements}

The authors thank Martin~H.~Gutknecht (ETH~Z\"{u}rich) for supervising the 2004 diploma
thesis \cite{schmelzer2004} on block MINRES and SYMMLQ, which motivated the present work.

\bibliographystyle{siamplain}
\bibliography{refs}

\end{document}
