\documentclass[10pt,twocolumn]{article}

\usepackage[margin=1.5cm, top=1.2cm, bottom=1.2cm, columnsep=0.6cm]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{booktabs}
\usepackage{hyperref}
\setlength{\parskip}{3pt}

\title{\textbf{Krylov Methods for the Minimum Variance Portfolio}\\[4pt]
       \large One-Page Summary}
\author{Thomas Schmelzer \and Michael Wolf \and Martin Stoll}
\date{}

\begin{document}
\maketitle
\thispagestyle{empty}

\subsubsection*{Problem}

The long-only minimum variance portfolio is conventionally solved by forming a sample covariance
matrix $\Sigma = X^\top X / T$ and passing it to a general-purpose QP solver. This paper argues
the covariance matrix is unnecessary: all required operations reduce to matrix-vector products on
the raw returns matrix $X \in \mathbb{R}^{T \times n}$.

\subsubsection*{Key Insight}

Working with $X$ directly, the equality-constrained problem yields a symmetric indefinite
saddle-point system of size $(n+1)\times(n+1)$:
\[
\begin{pmatrix} 2X^\top X & \mathbf{1} \\ \mathbf{1}^\top & 0 \end{pmatrix}
\begin{pmatrix} w \\ \lambda \end{pmatrix}
=
\begin{pmatrix} \mathbf{0} \\ 1 \end{pmatrix}.
\]
Standard CG requires positive definiteness and does not apply; MINRES \cite{paige1975} handles
the indefinite case and never forms $X^\top X$ explicitly. The long-only constraint $w \geq 0$
is handled by an active set loop.

\subsubsection*{Solvers Compared}

Four methods are benchmarked --- CVXPY/Clarabel (reference), direct KKT solve, MINRES, and
constraint-eliminated CG --- all producing identical weights. The comparison is purely
computational.

\subsubsection*{Ledoit-Wolf Shrinkage as Preconditioner}

Replacing $X$ with $\tilde{X} = [\sqrt{c}\,X;\,\sqrt{\gamma}\,I]$ incorporates Ledoit-Wolf
shrinkage \cite{ledoit2004} without forming $\Sigma_{\mathrm{LW}}$. This compresses the eigenvalue
spectrum and reduces Krylov iteration counts directly. Statistical regularisation and numerical
preconditioning act on the same object: the eigenvalue distribution of the system matrix.

\subsubsection*{Empirical Results}

\begin{center}
\small
\begin{tabular}{lrr}
\toprule
 & Synthetic & S\&P~500 \\
Method & speedup & speedup \\
\midrule
CVXPY (Clarabel) &   1$\times$ &   1$\times$ \\
KKT direct       &  98$\times$ & 117$\times$ \\
MINRES           & 644$\times$ & 354$\times$ \\
CG (elim.)       & 558$\times$ & 385$\times$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent Results shown with Ledoit-Wolf shrinkage applied. On S\&P~500 data, MINRES drops from
214 to 105 iterations with shrinkage; CG drops from 387 to 117. Real equity data is harder than
synthetic due to a dominant market factor that inflates the condition number, making shrinkage
more impactful in practice.

\subsubsection*{Scaling}

MINRES scales empirically as $O(n^2)$ vs.\ $O(n^3)$ for the direct solve; the gap widens beyond
$n \approx 300$. In the rank-deficient regime ($T < n$), shrinkage is essential: without it the
KKT system is singular and iterative solvers do not converge.

\subsubsection*{Conclusion}

The minimum variance portfolio problem, formulated directly in terms of the returns matrix, is a
structured linear algebra problem. Ledoit-Wolf shrinkage and Krylov preconditioning theory address
different problems but act on the same object. The KKT saddle-point structure extends naturally to
the full Markowitz mean-variance setting; this generalisation is developed in a companion paper.

\renewcommand{\refname}{\small References}
\begingroup
\small
\bibliographystyle{plain}
\bibliography{refs}
\endgroup

\end{document}
